\nchapter{Changelog}
% Changelog.  Version, date, comments
\newcommand{\changelog}[3]{\noindent • \textbf{#1} (#2): #3\\}

\noindent Recent and notable updates:
\medskip

% Don't forget to update the title on the front page for a version change.

{\small
\changelog{1.32}{Jul 10 2023}{usage of \N{fì'u} vs.\ \N{tsa'u}
nominalizations; \N{aylahe} short forms; more \N{pak} examples, and a
clarification; genitives of \N{po} derived pronouns (\N{frapo},
etc.); \N{nì-} is productive with pronouns; colloquial compounds
in \N{-tseng}; PRO-drop section merged with subject dropping section;
rewrote entire Constituent Order section; the usual collection of typo
fixes, minor style changes, and additional citations and
interlinears.}
\changelog{1.31}{Jan 22 2023}{more examples in lenition section, rather
than just links to sections; topical with \N{srake}; when to use
subjunctive with \N{txo}; note on genitives of \N{pesu, tupe};
adjective pile-ups and adjective + relative order; clarify elision of
final \N{-e}; can't drop \N{'efu} colloquially; ceremonial pronoun
updates (\N{oheng, poho}, etc); \N{ye'rìn} redundant with proximal
future; clarify productivity of \N{nì-}; more interlinearized
examples; subject dropping; Reef Na'vi dialect.}
\changelog{1.30}{Oct 12 2022}{new font; \N{-tu} clarifications;
colloquial omissions; new section, Semantics, which has the copula,
and perception; as always, more citation links; explain interlinears
and introduce for some examples.}
\changelog{1.29}{Feb 7 2022}{update CA license; declension of foreign
words; reflexive of causative; \N{tìk} as conjunction; minor style
changes and more links.} 
\changelog{1.28}{Jan 1 2021}{slang section added; many minor style and
formatting tweaks; a good dose of example, citation, and indexing
backfill; full set of \N{tsa-} conjunction contractions; visual
version of legal syllable rules; productivity of \N{-yu}
with \N{si}-verbs; clarification on omission of \N{san} or \N{sìk};
refinement on use of \N{pum}; numbers after \N{zam}; \N{-tsim}
compound element; modals that take \N{futa} section redone; syntax
section for \N{sno} redone; sequential verbs with modal; dative
in \N{-ru} with glottal stop}
\changelog{1.27}{Mar 5 2020}{Edited section on \N{tsnì} to reflect new information.}
\changelog{1.26}{Oct 21 2019}{added new \N{fu} subsubsection for choice
questions; fixed two counts of typo for word ``conjunction". Added new
subsection for lenition of proper nouns.}
\changelog{1.25}{}{negative polar questions; vocative location
possibilities; \N{sunkesun} specifics; model syntax with \N{ke}; new
modal verb} 
\changelog{1.24}{Jan 26 2018}{renamed ``Punch'' section ``Focus,'' and added
examples to it; notes about modal word order; reorganized ``Register''
section; various small formatting and style tweaks; example of \N{fu}
with verb phrases; clarify forms of \N{tsa'u, tsaw}; Plumps' reported
clarification on \N{sì}; firm statement from Paul about only 3rd.\ person
antecedents for \N{sno}.}
\changelog{1.23}{Dec 17 2017}{additional example sentences in the
Syntax chapter; tidied section on \N{-eyä} genitives; meanings of individual
adpositions removed --- we have good dictionaries for that, and I've
added links.}
\changelog{1.22}{Nov 28 2017}{More citations; \N{fìay+} vs.\ \N{fay+};
\N{fray+} confirmed; difference between \N{nì-} and \N{na/pxel} with
pronouns; \N{nì-} with ordinals; animals and \N{po}; \N{mì} in time
expressions detail.}
\changelog{1.21}{June 2 2013}{note modal syntax of \N{sto}; \N{kxamlä}
example; clarification on diphthong declension; ADJ$_i$-a N a-ADJ$_i$
construction; \N{-nay} derivation; \N{frafya}; free choice indefinites
with \N{ketsran}; contrafactuals, \N{zun} and \N{zel}.}
\changelog{1.20}{Nov 29 2012}{new meaning of \N{ìlä}; \N{akum/kuma}
with \N{nìftxan} and \N{fìtxan}; general condition; clearer
explanation of \N{srake}; note pronunciation of \N{säpoli} in casual
conversation; ``Affect and Evidence'' section; say more about pronoun
genitives.}
\changelog{1.19}{April 7 2012}{note temporal use of \N{mì}; added a
section on poetic register; new adp.\ \N{sko}; \N{sna-} and \N{-tswo}
compound elements; clarify infix position in V-V compounds; \N{krra}
not \N{krr a}; some detail on relative clauses and transitivity;
rename ``antipassive'' section to ``ambitransitivity,'' and add some
more details from blog post; as always, some more citations.}
\changelog{1.18}{March 4 2012}{\N{fìtsap} with intransitives;
intransitive reflexives; a bunch of typo fixes from LN.org member Blue
Elf; contrastive demonstratives; new info on \N{sä-}; correlative
comparisons; clean up redup\-li\-ca\-tion section, with help from
\N{nì'ul'ul} example.}
\changelog{1.17}{November 3 2011}{widened margins a bit and added a
toc; reciprocal w/ \N{fìtsap}; conjectural quest\-ions; \N{kam},
\N{kay}, \N{nuä}; more citations.}
\changelog{1.08}{October 15 2010}{Everything from the Language
Workshop\footnote{\texttt{http://wiki.learnnavi.org/index.php/Canon/2010/UltxaAyharyuä}}}
\changelog{1.0}{Aug 13 2010}{Initial release.}
}


